A Research on Big Data and AI Analysis Algorithm Optimization Using GPUs

### 2018-05-14 paper.tex
remove dicomo ref
update affi
ipa font
pos fixed
globally resize imgs

%\affiliate{IPSJ}{Information Processing Society of Japan}
\paffiliate{DICOMO}{DICOMO2018}
\affiliate{SOICT}{University of Tokyo, Graduate School of Information Science and Technology, Social ICT Center}
\affiliate{CS}{University of Tokyo, Graduate School of Information Science and Technology, Department of Information Science}

\author{TRAN VAN SANG}{CS}
\author{KOBAYASHI RYOUSUKE}{SOICT}
\author{YAMAGUCHI RIE}{SOICT}
\author{NAKATA TOSHIYUKI}{SOICT}

@@

For supervised learning model training purpose, the next step in the data process is negative sample generation. User id and weekday were uniformly randomized from their domains. All newly generated values' label are false. However, with uniform generator for remaining 24 features of number of reading, Decision Tree model produced a predictor that can correctly predict 99.31\% test samples, which implies a bad data generation rather than leveraging the training model's performance. Hence, we investigated new way to generate number of reading to generate data more likely to existing data. Because all users usually do not read manga on every hour, there should be more zero value in these 24 features. In order to do this, firstly, number of total reading of each user was investigated. Figure~\ref{fig:total_hist} 2 shows this value histogram.\newline

@@

@@

  \includegraphics[width=\columnwidth,natwidth=1024,natheight=295]{total_hist.png}

@@

  \vspace*{-2cm}

@@

  \includegraphics[width=\columnwidth,natwidth=2060,natheight=1398]{gap_stat.png}

@@

  \vspace*{-2cm}

@@

  \includegraphics[width=\columnwidth,natwidth=1024,natheight=805]{knn.png}

@@

  \vspace*{-2cm}

@@

  \includegraphics[width=\columnwidth,natwidth=1024,natheight=485]{dtree.png}

@@

@@

  \includegraphics[width=\columnwidth,natwidth=1021,natheight=151]{profvis.png}

@@

  \vspace*{-2.5cm}

@@

  \includegraphics[width=\columnwidth,natwidth=411,natheight=362]{wrapping.png}

@@

  \vspace*{-3.6cm}

@@

  \includegraphics[width=\columnwidth,natwidth=451,natheight=548]{qr_flow.png}

@@

  \includegraphics[width=\columnwidth,natwidth=827,natheight=805]{glm_coef.png}

@@

\begin{figure}[ht]
  \vspace*{-2.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=827,natheight=805]{access_var.png}
  \caption{Access count variance each hour in a day}\label{fig:access_var}
\end{figure}
\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=827,natheight=484]{access_mean.png}
  \caption{Mean number of reading by hour in a day}\label{fig:access_mean}
\end{figure}
\section{Optimization}
Time for training all data with standard R algorithm, manual LS solver and CUSOLVER LS solver were 112.79 seconds, 87.579 seconds, 88.897 seconds respectively. These timing result is described in detail in Figure~\ref{fig:qr_ret}\newline

\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{qr_ret.png}
  \caption{Training time of 3 implementation3}\label{fig:qr_ret}
\end{figure}
Figure~\ref{fig:qr_ret} shows that CUSOLVER LS solver C extension run 2.12 times faster than R standard version \texttt{C\_Cdqrls}. Because of the major portion of overhead timing, the overall algorithm can speed up merely 1.27 times over maximal ideal speed up rate of 2.2 times. C extension running time of CUSOLVER LS solver is slightly faster than manual LS solver's one. This can be explained by the fact that input matrix is indeed linearly independent, CUSOLVER LS solver implementation only need to run QR decomposition once without caring about linear independence while manual LS solver version does care about. In CUSOLVER LS solver running time, there exists portion named \texttt{R-C protocol copying overhead}. This is time occupied by memory transfer between R program and C extension because CUSOLVER LS solver program does not use optimized interface while manual LS solver does. This portion elimination can be adopted with not much effort.\newline

\begin{figure}[ht]
  \vspace*{-2cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=857,natheight=484]{qr_blas_ret.png}
  \caption{Relation between BLAS accelerated run time with number of OMP cores}\label{fig:qr_blas_ret}
\end{figure}

@@

\begin{figure}[ht]
  \vspace*{-1cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1902,natheight=550]{htop.png}
  \caption{htop catches a glimpse when all CPUs are fully used}\label{fig:htop}
\end{figure}

\begin{figure}[ht]
  \vspace*{-2cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{nvprof_qr.png}
  \caption{GPU profiling result via nvprof of CUSOLVER LS solver}\label{fig:nvprof_qr}
\end{figure}

\begin{figure}[ht]
  \vspace*{-2cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{nvprof_manual.png}
  \caption{GPU profiling result via nvprof of manual LS solver}\label{fig:nvprof_manual}
\end{figure}
From Figure~\ref{fig:nvprof_qr} and Figure~\ref{fig:nvprof_manual}, we can observe that CUDA memcpy routine, a memory transfer routine between CPU (host) and GPU (device), takes around 20\% of the execution on GPU device. This proves the potential of time reduction in future because in current implementation, all data is copied back to CPU after each iteration for convergence check while actually only a small portion of data is required.\newline

\begin{figure}[ht]
  \vspace*{-2.2cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{c_square.png}
  \caption{Isolated C extension time comparison on 3000 rows by 2000 columns matrix}\label{fig:c_square}
\end{figure}

\begin{figure}[ht]
  \vspace*{-1.5cm}
  \centering
  \includegraphics[width=\columnwidth,natwidth=1200,natheight=742]{c_narrow.png}
  \caption{Isolated C extension time comparison on 9,064,492 rows by 27 columns matrix}\label{fig:c_narrow}
\end{figure}
Figure~\ref{fig:c_square} and Figure~\ref{fig:c_narrow} emphasize matrix dimension's effect on C extension performance. With input matrix of 3,000 rows by 2,000 columns, GPU version speeded up LS solver by 17.94 times and 79.91 times resepectively. While with input matrix of 9,064,492 rows by 27 columns their speedup rate were merely 1.48 times and 1.65 times respectively. This implies that narrow matrix does not take more advantage of GPU speedup than square matrix.
\section{Conclusion}
Comparing with R. Kobayashi’s result in~\cite{kobayashi1}. General Linear model in this paper decreases both False Acceptance Rate and False Rejection Rate by 1.14 times (from 12.15\% down to 10.69\%) and 1.3 times (from 14.0\% down to 10.60\%) respectively, although these results slightly depends on model choice and false sample generation method. As discussed in methodology section, graph based approach is very promising. If it is well developed, it will not only be helpful on user classification and distinguishment but also has potential on content recommendation system. For example: user next reading chapter suggestion predictor.\newline
Due to major timing fraction of overhead timing, the optimized version of training algorithm achieved relatively high speedup of 1.27 times under 2.2 times upper bound limit. This speedup can be improved much more with convergence check improvement by reducing almost of memory transfer between GPU device and CPU host in middle iterations. The memory transfer should be done only in the first and the last iteration. Last experiment of the research proved that CUDA speedup effect highly depends on input matrix shape. That is to say, the more features input model has the more speedup this research's optimization gains. It is significant to take notice that both Least Squares solver implementations were not fully optimized. CUSOLVER LS solver requires calling QR decomposition twice due to CUDA Toolkit CUSOLVER library's low abstraction. On the other hand, manual LS solver is nothing more than a simple linear QR decomposition with Householder reflection. Blocked Householder QR algorithm in~\cite{Kerr:2009:QDG:1513895.1513904} is one of the fastest QR implementations executing entirely on the GPU~\cite{Kerr:2009:QDG:1513895.1513904}, with some more investigation on capability with linearly dependent system, it is clearly applicable and gains much speedup. In contrast to GPU optimization technique, it is worthwhile to take attention on multicore optimization by installing and loading appropriate lapack library to exploit multicore power.\newline
The last benchmark result on isolated C extension performance proves GPU accelerated algorithm's effectiveness to wide matrix. That is to say the accelerated algorithm gains more speedup on model with high number of features. Especially, when training model with categorical variables expanded, number of features increases proportionally to categorical variables' domain.

>>>>

%\affiliate{IPSJ}{Information Processing Society of Japan}
%\paffiliate{DICOMO}{DICOMO2018}
\affiliate{SOICT}{University of Tokyo, Graduate School of Information Science and Technology, Social ICT Center}
\affiliate{CS}{University of Tokyo, Graduate School of Information Science and Technology, Department of Computer Science}

\author{TRAN VAN SANG}{CS}
\author{KOBAYASHI RYOUSUKE}{SOICT}
\author{YAMAGUCHI RIE}{SOICT}
\author{NAKATA TOSHIYUKI}{SOICT,CS}

@@

For supervised learning model training purpose, the next step in the data process is negative sample generation. User id and weekday were uniformly randomized from their domains. All newly generated values' label are false. However, with uniform generator for remaining 24 features of number of reading, Decision Tree model produced a predictor that can correctly predict 99.31\% test samples, which implies a bad data generation rather than leveraging the training model's performance. Hence, we investigated new way to generate number of reading to generate data more likely to existing data. Because all users usually do not read manga on every hour, there should be more zero value in these 24 features. In order to do this, firstly, number of total reading of each user was investigated. Figure~\ref{fig:total_hist} shows this value histogram.\newline

@@

  \vspace*{-0.7cm}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1024,natheight=295]{total_hist.png}

@@

  \vspace*{-0.5cm}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=2060,natheight=1398]{gap_stat.png}

@@

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1024,natheight=805]{knn.png}

@@

  \vspace*{-0.7cm}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1024,natheight=485]{dtree.png}

@@

  \vspace*{-0.8cm}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1021,natheight=151]{profvis.png}

@@

  \vspace*{-0.5cm}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=411,natheight=362]{wrapping.png}

@@

  \vspace*{-0.7cm}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=451,natheight=548]{qr_flow.png}

@@

  \includegraphics[width=\columnwidth * 2 / 3,natwidth=827,natheight=805]{glm_coef.png}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=827,natheight=805]{access_var.png}
  \caption{Access count variance each hour in a day}\label{fig:access_var}
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=827,natheight=484]{access_mean.png}
  \caption{Mean number of reading by hour in a day}\label{fig:access_mean}
\end{figure}
\section{Optimization}
Time for training all data with standard R algorithm, manual LS solver and CUSOLVER LS solver were 112.79 seconds, 87.579 seconds, 88.897 seconds respectively. These timing result is described in detail in Figure~\ref{fig:qr_ret}\newline

\begin{figure}[ht]
  \vspace*{-0.8cm}
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1200,natheight=742]{qr_ret.png}
  \caption{Training time of 3 implementation3}\label{fig:qr_ret}
\end{figure}
Figure~\ref{fig:qr_ret} shows that CUSOLVER LS solver C extension run 2.12 times faster than R standard version \texttt{C\_Cdqrls}. Because of the major portion of overhead timing, the overall algorithm can speed up merely 1.27 times over maximal ideal speed up rate of 2.2 times. C extension running time of CUSOLVER LS solver is slightly faster than manual LS solver's one. This can be explained by the fact that input matrix is indeed linearly independent, CUSOLVER LS solver implementation only need to run QR decomposition once without caring about linear independence while manual LS solver version does care about. In CUSOLVER LS solver running time, there exists portion named \texttt{R-C protocol copying overhead}. This is time occupied by memory transfer between R program and C extension because CUSOLVER LS solver program does not use optimized interface while manual LS solver does. This portion elimination can be adopted with not much effort.\newline

\begin{figure}[ht]
  \vspace*{-0.8cm}
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=857,natheight=484]{qr_blas_ret.png}
  \caption{Relation between BLAS accelerated run time with number of OMP cores}\label{fig:qr_blas_ret}
\end{figure}

@@

\begin{figure}[ht]
  \vspace*{-0.6cm}
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1902,natheight=550]{htop.png}
  \caption{htop catches a glimpse when all CPUs are fully used}\label{fig:htop}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1200,natheight=742]{nvprof_qr.png}
  \caption{GPU profiling result via nvprof of CUSOLVER LS solver}\label{fig:nvprof_qr}
\end{figure}

\begin{figure}[ht]
  \vspace*{-0.2cm}
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1200,natheight=742]{nvprof_manual.png}
  \caption{GPU profiling result via nvprof of manual LS solver}\label{fig:nvprof_manual}
\end{figure}
From Figure~\ref{fig:nvprof_qr} and Figure~\ref{fig:nvprof_manual}, we can observe that CUDA memcpy routine, a memory transfer routine between CPU (host) and GPU (device), takes around 20\% of the execution on GPU device. This proves the potential of time reduction in future because in current implementation, all data is copied back to CPU after each iteration for convergence check while actually only a small portion of data is required.\newline

\begin{figure}[ht]
  \vspace*{-1.2cm}
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1200,natheight=742]{c_square.png}
  \caption{Isolated C extension time comparison on 3000 rows by 2000 columns matrix}\label{fig:c_square}
\end{figure}

\begin{figure}[ht]
  \vspace*{-0.5cm}
  \centering
  \includegraphics[width=\columnwidth * 2 / 3,natwidth=1200,natheight=742]{c_narrow.png}
  \caption{Isolated C extension time comparison on 9,064,492 rows by 27 columns matrix}\label{fig:c_narrow}
\end{figure}
Figure~\ref{fig:c_square} and Figure~\ref{fig:c_narrow} emphasize matrix dimension's effect on C extension performance. With input matrix of 3,000 rows by 2,000 columns, GPU version speeded up LS solver by 17.94 times and 79.91 times resepectively. While with input matrix of 9,064,492 rows by 27 columns their speedup rate were merely 1.48 times and 1.65 times respectively. This implies that narrow matrix does not take more advantage of GPU speedup than square matrix.
\section{Conclusion}
Comparing with R. Kobayashi's result in~\cite{kobayashi1}. General Linear model in this paper decreases both False Acceptance Rate and False Rejection Rate by 1.14 times (from 12.15\% down to 10.69\%) and 1.3 times (from 14.0\% down to 10.60\%) respectively, although these results slightly depends on model choice and false sample generation method. As discussed in methodology section, graph based approach is very promising. If it is well developed, it will not only be helpful on user classification and distinguishment but also has potential on content recommendation system. For example: user next reading chapter suggestion predictor.\newline
Due to major timing fraction of overhead timing, the optimized version of training algorithm achieved relatively high speedup of 1.27 times under 2.2 times upper bound limit. This speedup can be improved much more with convergence check improvement by reducing almost of memory transfer between GPU device and CPU host in middle iterations. The memory transfer should be done only in the first and the last iteration. Last experiment of the research proved that CUDA speedup effect highly depends on input matrix shape. That is to say, the more features input model has the more speedup this research's optimization gains. It is significant to take notice that both Least Squares solver implementations were not fully optimized. CUSOLVER LS solver requires calling QR decomposition twice due to CUDA Toolkit CUSOLVER library's low abstraction. On the other hand, manual LS solver is nothing more than a simple linear QR decomposition with Householder reflection. Blocked Householder QR algorithm in~\cite{Kerr:2009:QDG:1513895.1513904} is one of the fastest QR implementations executing entirely on the GPU~\cite{Kerr:2009:QDG:1513895.1513904}, with some more investigation on capability with linearly dependent system, it is clearly applicable and gains much speedup. In contrast to GPU optimization technique, it is worthwhile to take attention on multicore optimization by installing and loading appropriate lapack library to exploit multicore power.\newline
The last benchmark result on isolated C extension performance proves GPU accelerated algorithm's effectiveness to wide matrix. That is to say the accelerated algorithm gains more speedup on model with high number of features. Especially, when training model with categorical variables expanded, number of features increases proportionally to categorical variables' domain.
